Depression, Anxiety, and Objective Hardship: Examining Prenatal Mental Health During the Covid-19 Pandemic in Canada

### 2024-04-18 outputs/paper.qmd
improve measurement

---
title: "Depression, Anxiety, and Objective Hardship: Examining Prenatal Mental Health During the Covid-19 Pandemic in Canada"
author: 
  - Boxuan Yi
format: 
  pdf:
    documentclass: article
    geometry: margin = 1in
    abstract: "The outbreak of the Covid-19 pandemic was a substantial stressor, especially for pregnant individuals. Since depression and anxiety beyond moderate levels are considered clinically significant, this paper uses logistic regression to evaluate the probability of experiencing at least moderate levels of prenatal depression or anxiety in Canada during the pandemic. The analysis indicates that younger age, low levels of education and household income, and a high perception of danger to both oneself and the unborn are associated with higher levels of depression and anxiety."
date: "17 April 2024"
date-format: "D MMMM YYYY"
number-sections: true
thanks: "Code and data in this analysis is available at: https://github.com/Elaineyi1/Prenatal_Mental_Health"
bibliography: references.bib
toc: true
---
\newpage
# Introduction {#sec-introduction}
While pregnancy and childbirth are often moments of joy, they can also bring about distress as mothers navigate responsibilities and physical changes. In the 12 months before childbirth, 12.5% of women in Canada reported that most days were very stressful, and 7.5% reported depressive or anxious symptoms [@canadaPregnancyWomenaposs]. These figures were likely to increase during the COVID-19 pandemic due to heightened uncertainty and panic. Maternal depression and anxiety during pregnancy are associated with adverse neurodevelopmental outcomes for the offspring, such as hyperactivity disorder, autism, poor cognitive or emotional development [@MPMHPbeta], which emphasizes the importance of prenatal mental health for both mothers and children. In light of this, I utilize data from the Open Science Framework to explore factors contributing to maternal depression and anxiety during the pandemic, along with associated birth outcomes. This analysis could also provide insight into mothers’ mental health during other objective hardships or the general effects of prenatal distress.

@@

# Data {#sec-data}
## Data Overview and Measurement
The prenatal mental health dataset sourced from the Open Science Framework was designed to investigate the relationship between exposure to objective hardship caused by the Covid-19 pandemic and psychological distress among pregnant individuals [@osfPregnancyDuring]. The population is pregnant individuals aged 17 or older, with gestation periods not exceeding 35 weeks, residing in Canada. Data collection spanned from May 2020 to December 2021, containing individual-level demographic information, mental health data and basic birth outcomes [@nihPrenatalMental]. For this paper, I will only use data collected with delivery month from May to December in 2020 to analyze the prenatal mental health during the initial phase of the outbreak, a period highly marked by panic and adversity. 
Participants were recruited through pregnancy organizations, care providers, social media, and paid ads on Facebook and Instagram, with the chance to win a 500-dollar gift card. Some of the advertisements were specifically targeted to geographic regions and sociodemographic groups with lower representation to reduce underrepresentation [@nihPrenatalMental]. Surveys were available in either English or French. Prenatal data, including age, household income in Canadian dollars for the year 2019, and education level were collected as part of the national Pregnancy during the COVID-19 Pandemic (PdP) project using online questionnaires through REDCap. Birth data were also acquired using parents reports [@nihPrenatalMental].

@@

## Birth Weight/Length Related to Prenatal Mental Health
Studies have shown that maternal distress may contribute to reduced birthweight, and complications related to preterm birth and low birth weight are leading causes of infant morbidity and mortality [@NOWWUADCWWWD]. The dataset used in this paper, from the Open Science Framework [@osfPregnancyDuring], also demonstrates a correlation between poor mental health and reduced birth weight and birth length. As illustrated in [@fig-weight_length_delivery], there is a clear association between higher levels of depression or anxiety and lower birth weight and length, categorized by delivery mode, with Caesarean-section babies having a larger average length and weight compared to vaginally delivered babies. These findings, once again, emphasize the importance of addressing maternal mental health during pregnancy to support fetal growth and development.

>>>>

---
title: "Depression, Anxiety, and Objective Hardship: Examining Prenatal Mental Health During the Covid-19 Pandemic in Canada"
author: 
  - Boxuan Yi
format: 
  pdf:
    documentclass: article
    geometry: margin = 1in
    abstract: "The outbreak of the Covid-19 pandemic was a substantial stressor, especially for pregnant individuals. Since depression and anxiety beyond moderate levels are considered clinically significant, this paper uses logistic regression to evaluate the probability of experiencing at least moderate levels of prenatal depression or anxiety in Canada during the pandemic. The analysis indicates that younger age, low levels of education and household income, and a high perception of danger to both oneself and the unborn are associated with higher levels of depression and anxiety."
date: "17 April 2024"
date-format: "D MMMM YYYY"
number-sections: true
thanks: "Code and data in this analysis is available at: https://github.com/Elaineyi1/Prenatal_Mental_Health"
bibliography: references.bib
toc: true
---
\newpage
# Introduction {#sec-introduction}
While pregnancy and childbirth are often moments of joy, they can also bring about distress as mothers navigate responsibilities and physical changes. In the 12 months before childbirth, 12.5% of women in Canada reported that most days were very stressful, and 7.5% reported depressive or anxious symptoms [@canadaPregnancyWomenaposs]. These figures were likely to increase during the COVID-19 pandemic due to heightened uncertainty and panic. Maternal depression and anxiety during pregnancy are associated with adverse neurodevelopmental outcomes for the offspring, such as hyperactivity disorder, autism, poor cognitive or emotional development [@MPMHPbeta], which emphasizes the importance of prenatal mental health for both mothers and children. In light of this, I utilize data collected from the University of Calgary and Alberta Children's Hospital Research Institute to explore factors contributing to maternal depression and anxiety during the pandemic, along with associated birth outcomes. This analysis could also provide insight into mothers’ mental health during other objective hardships or the general effects of prenatal distress.

@@

# Data {#sec-data}
## Data Overview and Measurement
The prenatal mental health dataset created by the University of Calgary and Alberta Children's Hospital Research Institute was designed to investigate the relationship between exposure to objective hardship caused by the Covid-19 pandemic and psychological distress among pregnant individuals [@osfPregnancyDuring]. The population is pregnant individuals aged 17 or older, with gestation periods not exceeding 35 weeks, residing in Canada. Data collection spanned from May 2020 to December 2021, containing individual-level demographic information, mental health data and basic birth outcomes [@nihPrenatalMental]. The dataset was downloaded from Open Science Framework. For this paper, I will only use data collected with delivery month from May to December in 2020 to analyze the prenatal mental health during the initial phase of the outbreak, a period highly marked by panic and adversity.

Participants were recruited through pregnancy organizations, care providers, social media, and paid ads on Facebook and Instagram, with the chance to win a 500-dollar gift card. Some of the advertisements were specifically targeted to geographic regions and sociodemographic groups with lower representation to reduce underrepresentation [@nihPrenatalMental]. Surveys were available in either English or French. Prenatal data, including age, household income in Canadian dollars for the year 2019, and education level were collected as part of the national Pregnancy during the COVID-19 Pandemic (PdP) project using online questionnaires through REDCap. Birth data were also acquired using parents reports [@nihPrenatalMental].

@@

## Birth Weight/Length Related to Prenatal Mental Health
Studies have shown that maternal distress may contribute to reduced birthweight, and complications related to preterm birth and low birth weight are leading causes of infant morbidity and mortality [@NOWWUADCWWWD]. The dataset used in this paper also demonstrates a correlation between poor mental health and reduced birth weight and birth length. As illustrated in [@fig-weight_length_delivery], there is a clear association between higher levels of depression or anxiety and lower birth weight and length, categorized by delivery mode, with Caesarean-section babies having a larger average length and weight compared to vaginally delivered babies. These findings, once again, emphasize the importance of addressing maternal mental health during pregnancy to support fetal growth and development.
